Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-02 index.qmd
section:issues,CaseV; process:write

Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*. Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution. The mode (or the mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$. Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$. Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli. This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in the other.

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$. The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$. As a result, the dichotomous outcome of this comparison almost certainly favors text B over text A.

@@

<!-- 5. The effects of correlation on the discriminal difference -->
Notably, the correlation between the discriminal processes, $\rho$, plays a pivotal role in determining comparison outcomes by shaping the distribution of the discriminal difference between the stimuli. Specifically, as the correlation increases, reflecting a stronger dependence of the judge's perception of quality in one stimulus on his perception of the other, the distribution of the discriminal difference narrows. This narrowing ultimately impacts the area under the curve that determines the comparison outcome and, consequently, the conclusions drawn from the comparison.

@@

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli. For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming that the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 4. What does the zero correlation between stimuli imply? and does it hold? -->
As outlined in the previous section, the assumption of *zero correlation between stimuli* suggests that during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @fig-correlation, $\rho=0$). Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 5. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between heterogeneous stimuli due to certain judges' biases that resist cancellation. Research on text quality indicates that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These aspects, which are likely neither equally weighted nor opposing, may influence judges' perceptions unevenly. For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. Ignoring these relevant additional "traits" can result in biases that resist cancellation, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346] and ultimately violating the assumption of zero correlation. While direct evidence for the specific mechanisms causing biases is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 6. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures in CJ data, the statistical handling of this extra source of dependency has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores, which generates additional statistical and measurement issues, as discussed in section @sec-theory-issue2.

<!-- 7. But what are the challenges when these assumptions are violated? -->
In any case, the psychometric and statistical literature strongly advises against ignoring relevant additional traits or overlooking clustering (grouping) structures that may introduce dependencies between stimuli. First, ignoring additional aspects pertinent to the trait of interest can cause a dimensional mismatch, which can lead to an overestimation of the precision (reliability) of the trait's measurement [@Hoyle_et_al_2023, pp. 340-341] or, worse, introduce bias into the measurement process [@Ackerman_1989]. Additionally, overlooking relevant hierarchical (or grouping) structures can exacerbate the inflation of reliability by reducing the accuracy of the model's parameters [@Hoyle_et_al_2023, pp. 482]. Taken together, these oversights undermine the reliability of the measurement process and, since validity cannot exist without reliability [@Perron_et_al_2015, pp. 2], ultimately compromise the validity of the results derived from it.

<!-- forcing a unidimensional trait structure on multidimensional data can inflate the perceived measurement precision (reliability) of the trait [@Hoyle_et_al_2023, pp. 340-341] -->

@@

<!-- 8. So, how do we mitigate these risks? -->
<!-- Luckily, @Andrich_1978 and @Wainer_et_al_1978 already proposed a solution to the first scenario where the assumption of zero correlation is violated. Both authors demonstrated that by parametrizing judges' biases in the measurement model of the trait of interest, their effects are experimentally eliminated from the trait's measurement process. Furthermore, a wealth of statistical literature offers a solution to the second scenario: incorporate the hierarchical structure into the measurement model of the trait of interest, similar to a multilevel SEM model [@Hoyle_et_al_2023, chap. 26]. -->

@@

>>>>

Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*. Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution [@Andrich_1978]. The mode (or the mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$. Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$. Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli. This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in the other.

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$. The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$. As a result, the dichotomous outcome of this comparison would mostly favor text B over text A.

@@

<!-- 5. The effects of correlation on the discriminal difference -->
Notably, the correlation between the discriminal processes, $\rho$, plays a pivotal role in determining comparison outcomes by shaping the distribution of the discriminal difference between the stimuli. Specifically, as the correlation increases, reflecting a stronger dependence of the judge's perception of quality in one stimulus on his perception of the other, the distribution of the discriminal difference narrows. This narrowing ultimately impacts the area under the curve that determines the comparison outcome and, consequently, the conclusions drawn from this outcome.

@@

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli. For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting them to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming that the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 4. What does the zero correlation between stimuli imply? and does it hold? -->
As outlined in @sec-thurstone_theory, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 5. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluating these traits, they often rely on various intricate characteristics of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, consequently, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 6. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures in CJ data, the statistical handling of this extra source of dependency has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores, which generates additional statistical and measurement issues, as discussed in section @sec-theory-issue2.

<!-- 7. But what are the challenges when these assumptions are violated? and how to solve them? -->
In any case, the psychometric and statistical literature underscores the importance of incorporating additional relevant traits and clustering (grouping) structures into the modeling of CJ data to address potential dependencies between stimuli. Oversights in these areas can lead to misleading conclusions about the trait's reliability and ultimately compromise its validity [@Perron_et_al_2015, pp. 2]. Neglecting additional traits relevant to the stimuli or the trait of interest, such as judges' biases, introduces a dimensional mismatch in the modeling process of the CJ data. This mismatch often results in an overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 341] or, worse, introduces biases into the estimated trait [@Ackerman_1989]. Therefore, these traits must be incorporated into the modeling of CJ data, as demonstrated by @Andrich_1978 and @Wainer_et_al_1978. Similarly, ignoring hierarchical (grouping) structures reduces the accuracy of model parameter estimates and further exacerbates the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. To address this, any additional hierarchical structure relevant to the stimuli should also be included in the model, as in Multilevel Structural Equation models (MSEM) [@Hoyle_et_al_2023, chap. 26].

<!-- This further implies that the shape of the discriminal difference distribution that determines the comparison outcome and, consequently, the conclusions drawn from this outcome (see @sec-thurstone_theory, @fig-correlation). -->

<!-- To show the practical consequences, in Figure 18.1, we present the test information for the two models; clearly, the information is positively biased when the multidimensional data are forced into a unidimensional IRT model. Moreover, in this same figure, we show the test response curves (how expected scores change as a function of trait level) for the two models; again, they, clearly, are not overlapping. The estimated implication is that if one applied a unidimensional IRT model, psychometric information is inflated due to the multidimensionality and examinee standard errors, which are the inverse square root of information, are too small. [@Hoyle_et_al_2023, pp. 340-341] -->

<!-- Accounting for measurement error in defining latent variables improves the accuracy of the model’s estimated structural parameters between the latent constructs, which in turn can provide a stronger test of a proposed theoretical model. [@Hoyle_et_al_2023, pp. 482] -->

@@

@@

<!-- The correlation $\rho$, considered above is not the kind of correlation considered by Bock (1958) in studies of preference and choice of objects. The Bock correlation arises from a specific interaction between a subject and an item because the subject has a particular "preference", such as a taste preference, for the object. This preference for the object is then reflected in its comparisons with all other objects. However, it is less likely that subjects would have such particular preferences for some items in a test and, in any case, the assumption is open to empirical checking [@Andrich_1978, pp. 455]. -->
